---
title: "ggplot2 -- The grammar of graphics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```

```{r, echo=FALSE, fig.width=6, fig.height=4.5}
library(palmerpenguins)
library(ggplot2)

theme_new = function() {  # this for all the elements common across plots
  theme_void() %+replace%
    theme(panel.border = element_rect(color="black",size=0.5, fill = NA),
          plot.margin = margin(0.5, 0.5, 0.5, 0.5, "mm"))
}

point = ggplot(penguins, aes(x = body_mass_g, flipper_length_mm)) + 
    geom_point(size = 0.5) + 
    theme_new() + labs(title = "scatterplot") +
    annotate("text", label = "+ geom_point()", x = 2500, y = 225, hjust = "left", vjust = "top", size = 3)

smooth = ggplot(penguins, aes(x = body_mass_g, flipper_length_mm))+
  geom_smooth(color = "black") +
  theme_new() + labs(title = "smoother") +
  annotate("text", label = "+ geom_smooth()", x = 2500, y = 225, hjust = "left", vjust = "top", size = 3)


line = ggplot(penguins, aes(x = body_mass_g, flipper_length_mm)) + 
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_new() + labs(title = "linear regression")+
  annotate("text", label = "+ geom_smooth(\nmethod = *lm*)", x = 2500, y = 225, hjust = "left", vjust = "top", size = 3)


jitter = ggplot(penguins, aes(x = species, flipper_length_mm)) +
  geom_jitter(width = 0.1) + 
  theme_new() + labs(title = "jitter plot")+
  annotate("text", label = "+ geom_jitter()", x = 0.5, y = 225, hjust = "left", vjust = "top", size = 3)


bar = ggplot(penguins, aes(x = species, flipper_length_mm)) + 
  geom_bar(stat = "identity") +
  theme_new() + labs(title = "bar plot")+
  annotate("text", label = "+ geom_bar(stat = *identity*)", x = 0.5, y = 32000, hjust = "left", vjust = "top", size = 3)


boxplot = ggplot(penguins, aes(x = species, flipper_length_mm)) + 
  geom_boxplot() +
  theme_new() + labs(title = "boxplot")+
  annotate("text", label = "+ geom_boxplot()", x = 0.5, y = 225, hjust = "left", vjust = "top", size = 3)


violin = ggplot(penguins, aes(x = species, flipper_length_mm)) + 
  geom_violin() +
  theme_new() + labs(title = "violin plot")+
  annotate("text", label = "+ geom_violin()", x = 0.5, y = 225, hjust = "left", vjust = "top", size = 3)


histogram = ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram() +
  theme_new() + labs(title = "histogram")+
  annotate("text", label = "+ geom_histogram()", x = 170, y = 35, hjust = "left", vjust = "top", size = 3)


density = ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_density() +
  theme_new() + labs(title = "density plot")+
  annotate("text", label = "+ geom_density()", x = 170, y = 0.04, hjust = "left", vjust = "top", size = 3)


# dotplot = ggplot(penguins, aes(x = species, y = flipper_length_mm))+
#   geom_dotplot(binaxis = "y", dotsize = 0.5, stackdir = "center") +
#   theme_void()+theme(panel.border = element_rect(fill = NA))

library(patchwork)
# (point + smooth + line) / (bar + boxplot + violin + jitter) / (histogram + density + 
# plot_spacer()+theme_void())



cowplot::plot_grid(point, smooth, line, plot_spacer, bar, boxplot, violin, jitter, histogram, density, nrow = 3)


```

----

This tutorial uses the `{palmerpenguins}` package, available on CRAN.

```{r}
library(palmerpenguins)
library(ggplot2)
```

first, use the `str` function for a quick exploration of columns and data types
```{r}
str(penguins)
```

----

## Elements of a `ggplot` graph: {#toc}

1. [foundation](#foundation)
1. [`geoms`](#geoms)
1. [formatting `geom`s](#formatting)
1. [color and shape scales](#scales)
1. [formatting axes](#axes)
1. [labels and annotations](#labs)
1. [themes](#themes)

****

### Step 1: set up the basic foundation plot {#foundation}
[*back to top*](#toc)

Specify (a) the dataset being used and (b) the mapping (x and y axes)

`aes` stands for *aesthetic*, which is used to set axes and grouping variables (see below)

```{r, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm))
```

It is not necessary to use `data = ` and `mapping = `.

```{r}
ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm))
```

---

### Step 2: add `geoms` {#geoms}
[*back to top*](#toc)

`ggplot2` uses `geom` functions to add layers with different types of data visualization, e.g. points, lines, boxplots, violin plots, etc. Simply add `+` at the end of the foundation line and add new layers as needed.

**example: scatterplot**
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()
```

you can overlay multiple `geoms`

**example: scatterplot with linear regression line**
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

**example: boxplot with points**
```{r}
ggplot(penguins, aes(x = island, y = flipper_length_mm)) +
  geom_boxplot()+
  geom_point()
```

---

### Step 3: format `geoms` {#formatting}
[*back to top*](#toc)

Modify color, shape, size by adding arguments in the `geom_()` parentheses.  
If you leave the parentheses blank, it will default to `color = "black"`, `shape = 19`, `size = 1.5` for scatterplots

#### setting one color/shape for all points
simply type `color = ...`, etc. arguments in the `geom_()` parentheses. 

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(color = "red", shape = 19, size = 2)
```

Notes:

1. you can use "color" or "colour"
1. you can specify colors using the hex code (e.g. #000000) or color names from [this list](http://sape.inf.usi.ch/quick-reference/ggplot2/colour).
1. you can specify shapes using codes from [this list](http://www.sthda.com/english/wiki/ggplot2-point-shapes).


for `geom_smooth()` and similar line-based `geoms`, `size = ...` will set the line thickness
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue", size = 2)
```

for `geom_smooth()` and similar line-based `geoms`, `linetype = ...` will set the dash type. Options include `solid`, `dashed`, `longdash`, `dotted`, `dotdash` 
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_smooth(method = "lm", se = FALSE, linetype = "longdash", size = 2)
```



#### setting colors/shapes by group

use `aes(color = GROUPING VARIABLE, shape = GROUPING VARIABLE)` arguments in the `geom_()` parentheses. 

This example assigns a different color and shape for each island.
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = island, shape = island))
```

This example assigns a different color and dot size for each island.
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = island, size = island))
```

In the examples above, the grouping variables were categorical.  
You can also use numeric variables - you will get a gradient of color/size 

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(size = body_mass_g))
```

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = body_mass_g))
```


#### setting colors/shapes for multiple `geoms`

You can use the technique above for each `geom`. Use this to customize each `geom` separately.
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(color = "lightblue", size = 3)+
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 2)
```

Similarly, you can set color by group for each `geom`. 

In this example, we set color by species for the scatterplot only, and keep the regression line black.
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 2)
```

In this example, we set color by species for both, scatterplot and regression line.
```{r, fig.height=3, fig.width=4.5, eval=FALSE}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species))
```

We repeated the `aes(color = species)` for both `geoms`.  
A more efficient way is to add this argument to the foundation plot line.

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

What if you want separate regression lines, but all the same color? Use `aes(group = ...)`
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", se = FALSE, aes(group = species), color = "black")
```


Note: if there are contradicting pieces of code, the most recent occurrence will override the previous one.

In this example, we want to set color by species (`aes(color = species`)), but we also have `geom_point(color = "blue")`, which will override the previous color argument.
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, aes(color = species))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, aes(group = species), color = "black")
```


#### *color* vs. *fill*

`color` refers to outline color, whereas `fill` refers to object fill color.

`fill` is typically used for violin plots, boxplots, barplots. 

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_violin(color = "black", fill = "yellow")
```

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_violin(color = "black", aes(fill = species))
```

`fill` is can also be applied to `geom_point()` with certain shapes 
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(shape = 21, size = 2, color = "black", aes(fill = species))
```


#### transparency
Set transparency/opacity for `fill` using `alpha = ...`. `alpha = 0` is transparent and `alpha = 1` is opaque.

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_violin(color = "black", fill = "yellow", alpha = 0.3)
```




---

### Step 4: scales for colors and shapes {#scales}
[*back to top*](#toc)

**Note: These `scale_` functions work only if you use the `aes(color = ...)` or `aes(shape = ...)` functions to group by color/shape.**

    
#### manual color scales (for categorical variables)

You can set your own color palettes by using `scale_color_manual()`.  
Specify the colors you want using the [color names](http://sape.inf.usi.ch/quick-reference/ggplot2/colour) or hex codes. 

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species))+
  scale_color_manual(values = c("red", "blue", "yellow"))
```

You can use the [`brewer` color palettes](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/) 
```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species))+
  scale_color_brewer(palette = "Dark2")
```

You can also use other pre-defined palettes created by other users. Included here are [`{PNWPalettes}`](https://github.com/jakelawlor/PNWColors) and [`{soilpalettes}`](https://github.com/kaizadp/soilpalettes).

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species))+
  scale_color_manual(values = PNWColors::pnw_palette("Sunset2",3))
```

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species))+
  scale_color_manual(values = soilpalettes::soil_palette("podzol",3))
```

#### gradient color scales (for numeric/continuous variables)

When setting colors for a continuous variable, use `scale_color_gradient*()`.

Use `scale_color_gradient()` for a two-color gradient. You can set the high and low extremes.

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = body_mass_g))+
  scale_color_gradient(low = "red", high = "blue")
```


Use `scale_color_gradient2()` for a diverging color gradient. You can set the high and low extremes as well as the midpoint and midpoint color.

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = body_mass_g))+
  scale_color_gradient2(low = "red", high = "blue", mid = "white", midpoint = 4000)
```


Use `scale_color_gradientn()` if you want to use multiple colors in your gradient. You can also use a pre-defined color palette, e.g. `{soilpalettes}`.

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = body_mass_g))+
  scale_color_gradientn(colors = c("blue", "green", "yellow", "orange", "red"))
```

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = body_mass_g))+
  scale_color_gradientn(colors = soilpalettes::soil_palette("podzol", 5))
```


#### scales for shapes 
Similar to colors, you can set the shapes using `scale_shape_manual()`, using the codes [here](http://www.sthda.com/english/wiki/ggplot2-point-shapes).

```{r, fig.height=3, fig.width=4.5}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(shape = species))+
  scale_shape_manual(values = c(1, 16, 4))
```


 ---


### Step 5: formatting axes {#axes} 
[*back to top*](#toc)

#### setting axis limits

Option 1: `xlim()` and `ylim()`  
Option 2: `scale_x_continuous()` and `scale_y_continuous()`

Option 2 may look more complicated, but it is useful when you want to customize the axes even more (e.g. breaks and ticks)

```{r, eval = F}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()+
  xlim(3000, 5000)+
  ylim(130, 250)
```

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()+
  scale_x_continuous(limits = c(3000, 5000))+
  scale_y_continuous(limits = c(130, 250))
```

#### setting axis labels

Option 1: `xlab()` and `ylab()`  
Option 2: `labs(x = ..., y = ...)`

```{r, eval = F}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()+
  xlab("body mass (g)")+
  ylab("flipper length (mm)")
```

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()+
  labs(x = "body mass (g)", y = "flipper length (mm)")
```


<details>   
   <summary>for tips on more complicated axis labels, click here</summary>

coming soon. good luck.

</details>  

#### axis scales
You can use `scale_x_*()` and `scale_y_*` to set axis label, limits, major and minor breaks, and axis position

When working with a numerical/continuous axis, use `scale_x_continuous()`
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()+
  scale_x_continuous(limits = c(4000, 6000), 
                     breaks = seq(4000, 5000, by = 500), 
                     minor_breaks = seq(5000, 6000, by = 100))
```

When working with a categorical axis, use `scale_x_discrete()`

In this example, we set the x axis label and rename the tick labels (species).
```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) + 
  geom_point()+
  scale_x_discrete("Penguin Species", 
                   labels = c("Adelie" = "A", "Chinstrap" = "C", "Gentoo" = "G"))
```
 
```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_point()+
  scale_x_discrete(position = "top")
```


You can also modify the following aspects of axes, but those must be done through `theme()`:
- font size
- font family 
- bold, italics, etc.
- axis margin/spacing

---

### Step 6: themes {#theme} 
[*back to top*](#toc)




# ----

# ----


### line plots and smoothers
```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_smooth()
```

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_smooth(method = "lm")
```

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_smooth(method = "lm", se = FALSE)
```


### violin plots
```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_violin()
```


### boxplots
```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_boxplot()
```


### bar plots

Bar plots are used to represent the mean (and error), so those must be calculated before creating the plot.

```{r}
library(dplyr)
penguins_flipperlength_summary = 
  penguins %>% 
  group_by(species) %>% 
  dplyr::summarise(mean_flipper_length_mm = mean(flipper_length_mm, na.rm = TRUE),
                   sd = sd(flipper_length_mm, na.rm = TRUE),
                   se = sd/sqrt(n()))
```

```{r}
ggplot(penguins_flipperlength_summary, aes(x = species, y = mean_flipper_length_mm)) +
  geom_bar(stat = "identity")
```

```{r}
ggplot(penguins_flipperlength_summary, aes(x = species, y = mean_flipper_length_mm)) +
  geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin = mean_flipper_length_mm - sd, ymax = mean_flipper_length_mm + sd))
```



### distribution curves

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram()
```

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 100)
```


### combining geoms

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```



### stacked barplots

```{r}
penguins_count = 
  penguins %>% 
  mutate(count = 1)
```

```{r}
ggplot(penguins_count, aes(x = island, y = count))+
  geom_bar(stat = "identity", aes(fill = species))
```







