---
title: "tidyverse"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```


The `{tidyverse}` is a nice alternative to many base R functions for data cleaning and visualization.
`{tidyverse}` is a set of eight packages that perform different functions.

1. ggplot2
1. tibble
1. tidyr
1. readr
1. dplyr
1. purr
1. stringr
1. forcats

```{r}
library(tidyverse)
```

---

We will use the `{dplyr}` built-in dataset `starwars`

<details>
  <summary>click here to explore the `starwars` dataset</summary>

```{r}
names(starwars)
skimr::skim(starwars)

```

</details>

---

---

### A `{dplyr}` example (without pipes)
Consider this multi-part example: 

From the `starwars` dataset, you want to choose only characters with black hair. 

```{r}
sw_blackhair = filter(starwars, hair_color == "black")
sw_blackhair$name
names(sw_blackhair)
```

You then want to keep only columns related to name, height, and hair color

```{r}
sw_blackhair2 = select(sw_blackhair, name, height, hair_color)
names(sw_blackhair2)
```

You then want to rename the `hair_color` column

```{r}
sw_blackhair3 = rename(sw_blackhair2, "HairColor" = "hair_color")
names(sw_blackhair3)
```


In this example above, we created multiple files to get to our final desired output. An easier way to do this is using the **pipe** (%>%).

```{r}
sw2 = starwars %>% 
  filter(hair_color == "black") %>% 
  select(name, height, hair_color) %>% 
  rename("HairColor" = "hair_color")

names(sw2)
str(sw2)
```

---

---

# `{dplyr}` functions - the basics

## rename

```{r}
starwars %>% 
  rename("NAME" = "name") %>% 
  names()

starwars %>% 
  rename("NAME" = "name",
         "HEIGHT" = "height",
         "MASS" = "mass") %>% 
  names()
```

## select

These functions allow you to subset a dataset.

Use `select` to subset certain columns. You can do this multiple ways:

- choosing column names

```{r, eval=FALSE}
starwars %>% 
  select(name, height, mass) %>% 
  head()
```

- choosing column positions
```{r, eval=FALSE}
starwars %>% 
  select(1:3) %>% 
  head()
```

- deselecting column names/positions
```{r}
starwars %>% 
  select(-(4:14)) %>% 
  head()
```



## filter

Use `filter` to subset rows where certain conditions are met. 

For example, filter only characters with black hair 

```{r}
starwars %>% 
  filter(hair_color == "black") %>% 
  head()
```

#### multiple conditions 

For example, keep only characters with black or blond hair 
```{r}
starwars %>% 
  filter(hair_color %in% c("black", "blond")) %>% 
  head()
```
  

#### and/or conditions (multiple variables)
For example, keep only characters with black hair and blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" & eye_color == "blue") %>% 
  head()
```

keep only characters with black hair *or* blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" | eye_color == "blue") %>% 
  head()
```

#### "not" conditions

keep only characters without black hair 
```{r, eval=FALSE}
starwars %>% 
  filter(hair_color != "black") %>% 
  head()
```

another notation for this is 
```{r}
starwars %>% 
  filter(!hair_color == "black") %>% 
  head()
```


## mutate

### create a new column
### modify an existing column


## summarize

## group_by
group operations


## recode

## arrange

## drop_na


# `{dplyr}` functions - next level

## distinct

## starts_with and ends_with

## string contains

## join

## bind_rows and bind_columns

## if_else

## case_when

## pull



---

<details>
  <summary>Session Info</summary>

Date Run: `r Sys.Date()`

```{r}
sessionInfo()
```

</details>