---
title: "ggplot practice"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE)
```

Practice plots for data visualization/`{tidyverse}` skills. 

You may need to process the data files before you can create the plots. Hints and code are provided for each plot. 

---

## Palmer Penguins

Source: [`{palmerpenguins}`](https://github.com/allisonhorst/palmerpenguins). 
This dataset can be accessed by loading the `{palmerpenguins}` package, see below. This will load the file `penguins`.

```{r, eval=FALSE, echo=TRUE}
library(palmerpenguins)
```

This graph was inspired by [Cédric Scherer](https://github.com/Z3tt/TidyTuesday/tree/master/plots/2020_31)

```{r penguins_gg, fig.width=5, fig.height=5}
library(palmerpenguins)
library(tidyverse)
pal <- c("#FF8C00", "#A034F0", "#159090")


# plot 1 ----
gg_penguins1 =
  penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species, fill = species))+
  geom_point(aes(size = (body_mass_g/1000)), alpha = 0.2, shape = 21, stroke = 1)+
  scale_color_manual(values = pal)+
  scale_fill_manual(values = pal)+
  
  xlim(27,60)+
  labs(x = "bill length (mm)",
       y = "bill depth (mm)",
       size = "body mass (kg)",
       #title = "Exploring the PalmerPenguins Dataset",
       caption = "
       
       Source = PalmerPenguins") +
  
  annotate("text", label = "ADELIE", x = 34, y = 20, color = "#FF8C00", fontface = "bold")+
  annotate("text", label = "CHINSTRAP", x = 56, y = 19, color = "#A034F0", fontface = "bold")+
  annotate("text", label = "GENTOO", x = 55, y = 15, color = "#159090", fontface = "bold")+
  
  guides(color=FALSE, fill = FALSE, size=guide_legend(title.position = "top"))+
  scale_y_continuous(breaks = seq(12,20, by = 2))+
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = c(0.22, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = NA),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.caption = element_text(hjust = 0.5)
  )


# plot 2 ----

penguins_summary = 
  penguins %>% 
  group_by(species) %>% 
  dplyr::summarise(length_mean = mean(bill_length_mm, na.rm = TRUE),
                   length_median = median(bill_length_mm, na.rm = TRUE),
                   length_se = sd(bill_length_mm, na.rm = TRUE)/sqrt(n()),
                   length_sd = sd(bill_length_mm, na.rm = TRUE),
                   depth_mean = mean(bill_depth_mm, na.rm = TRUE),
                   depth_se = sd(bill_depth_mm, na.rm = TRUE)/sqrt(n()),
                   depth_sd = sd(bill_depth_mm, na.rm = TRUE)) %>% 
  left_join(penguins, by = "species")

gg_penguins1 + 
  geom_errorbar(data = penguins_summary, 
                aes(xmin = length_mean-length_sd, 
                    xmax = length_mean+length_sd, 
                    y = depth_mean,
                    color = species),
                width = 0.5, size = 1) +
  geom_errorbar(data = penguins_summary, 
                aes(ymin = depth_mean-depth_sd, 
                    ymax = depth_mean+depth_sd, 
                    x = length_mean,
                    color = species),
                width = 0.8, size = 1) +
  annotate("text", label = "error bars = mean ± standard deviation", x = 52, y = 12, size = 3)
```


<details>
<summary>Need a hint? Click here</summary>

- This graph uses two datasets: (a) the `penguins` dataset to create the scatterplot, and (b) a summary dataset for the error bars, which must be created from the `penguins` dataset.
- Use `{dplyr}` tools (`group_by()`, `summarize()`) to calculate means and standard deviations.
- Color palette used is: `pal <- c("#FF8C00", "#A034F0", "#159090")`.


</details>


<details>
<summary>Need the code? Click here</summary>

```{r penguins_code, eval = FALSE, echo = TRUE}
# load packages
library(palmerpenguins)
library(tidyverse)

# set the color palette
pal <- c("#FF8C00", "#A034F0", "#159090")


# plot 1 (base scatterplot) ----
gg_penguins1 =
  penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species, fill = species))+
  
  geom_point(aes(size = (body_mass_g/1000)), alpha = 0.2, shape = 21, stroke = 1)+
  
  # we are using shape = 21 (filled circle), which has both color and fill arguments
  scale_color_manual(values = pal)+
  scale_fill_manual(values = pal)+
  
  # set axis labels, title, caption, etc. 
  labs(x = "bill length (mm)",
       y = "bill depth (mm)",
       size = "body mass (kg)",
       #title = "Exploring the PalmerPenguins Dataset",
       caption = "
       
       Source = PalmerPenguins") +
  
  # add text via annotations to label the three species
  # this is an alternative to putting the colors in the legend
  annotate("text", label = "ADELIE", x = 34, y = 20, color = "#FF8C00", fontface = "bold")+
  annotate("text", label = "CHINSTRAP", x = 56, y = 19, color = "#A034F0", fontface = "bold")+
  annotate("text", label = "GENTOO", x = 55, y = 15, color = "#159090", fontface = "bold")+
  
  guides(color=FALSE, fill = FALSE, size=guide_legend(title.position = "top"))+
  
  # set axis limits
  # x axis is easier, but y-axis needs the scale_* function to set breaks
  xlim(27,60)+
  scale_y_continuous(breaks = seq(12,20, by = 2))+
  
  # set and customize the theme
  theme_bw()+
  theme(panel.grid = element_blank(),
        legend.position = c(0.22, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = NA),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.caption = element_text(hjust = 0.5)
  )


# plot 2 (error bars) ----

# create a summary dataframe that has mean/sd values
penguins_summary = 
  penguins %>% 
  group_by(species) %>% 
  dplyr::summarise(length_mean = mean(bill_length_mm, na.rm = TRUE),
                   length_sd = sd(bill_length_mm, na.rm = TRUE),
                   depth_mean = mean(bill_depth_mm, na.rm = TRUE),
                   depth_sd = sd(bill_depth_mm, na.rm = TRUE),
                   #length_se = sd(bill_length_mm, na.rm = TRUE)/sqrt(n()),
                   #depth_se = sd(bill_depth_mm, na.rm = TRUE)/sqrt(n())
  ) %>% 
  left_join(penguins, by = "species")


# now use this summary dataframe to add error bars onto the first plot
gg_penguins1 + 
  geom_errorbar(data = penguins_summary, 
                aes(xmin = length_mean-length_sd, 
                    xmax = length_mean+length_sd, 
                    y = depth_mean,
                    color = species),
                width = 0.5, size = 1) +
  geom_errorbar(data = penguins_summary, 
                aes(ymin = depth_mean-depth_sd, 
                    ymax = depth_mean+depth_sd, 
                    x = length_mean,
                    color = species),
                width = 0.8, size = 1) +
  annotate("text", label = "error bars = mean ± standard deviation", x = 52, y = 12, size = 3)
```

</details>



---

## The Big Mac Index
Source: [TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-12-22)

Use this code to download the data 

```{r bigmac_load, echo=TRUE}
big_mac <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-12-22/big-mac.csv')
```

```{r bigmac_gg, fig.width=10, fig.height=6}
library(tidyverse)

# data processing 

## process the data file, using usd_adjusted as reference
## create a new variable for over/under

big_mac_processed = 
  big_mac %>% 
  filter(!is.na(usd_adjusted)) %>% 
  mutate(valued = if_else(usd_adjusted > 0, "over", "under")) 


# data visualization -- 1

gg_bigmac_turkey =
  big_mac_processed %>% 
  filter(name == "Turkey") %>% 
  
  ggplot(aes(x = date, y = usd_adjusted)) +
  geom_hline(yintercept = 0, linetype = "longdash", alpha = 0.4)+
  geom_path(color = "darkred", size = 1)+
  geom_point(aes(shape = valued), color = "darkred", size = 3, stroke = 1, fill = "white") + 
  scale_shape_manual(values = c(21, 16))+
  
  labs(title = "The Big Mac Index: Turkey",
       subtitle = "The big mac index was invented by The Economist in 1986
       as a lighthearted guide to whether currencies are at their “correct” level.
       
       This graph shows how the Turkish Lira has compared to the US Dollar.",
       caption = "#TidyTuesday | Source: The Economist",
       x = "",
       y = "GDP adjusted index, 
       relative to the US dollar") + 
  
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 8, hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5),
    plot.margin = margin(t = 2, r = 1, b = 2, l = 1, unit = "cm")
  ) 



# data visualization -- 2

gg_bigmac_all = 
  big_mac_processed %>% 
  filter(name %in% c("Argentina", "Brazil", "Britain", 
                     "Canada", "Chile", "Egypt",
                     "Peru", "Poland", "Sweden")) %>% 
  
  ggplot(aes(x = date, y = usd_adjusted)) +
  geom_hline(yintercept = 0, linetype = "longdash", alpha = 0.4)+
  geom_path(color = "darkred")+
  geom_point(aes(shape = valued), color = "darkred", size = 1.5, fill = "white") + 
  scale_shape_manual(values = c(21, 16))+
  
  labs(title = "The Big Mac Index
       ",
       caption = "
       #TidyTuesday | Source: The Economist",
       x = "",
       y = "GDP adjusted index, 
       relative to the US dollar") + 
  facet_wrap(~name)+
  
  theme_void() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 8, hjust = 0.5),
    plot.caption = element_text(size = 8, hjust = 0.5),
    plot.margin = margin(t = 2, r = 1, b = 2, l = 1, unit = "cm")
  ) 


cowplot::plot_grid(gg_bigmac_turkey, gg_bigmac_all)
```

<details>
<summary>Need a hint? Click here</summary>

- We are setting different shapes for + and - values. To do this, create a new binary variable for above/below, and use that in `aes(shape = ...)`.
- Horizontal line at y = 0: `geom_hline (yintercept = 0)`
- Gridlines can be removed using `theme()`

For the second plot,  

- Choose a few countries using `filter(name %in% c(...))`
- Panels created using `facet_wrap()` 

</details>


<details>
<summary>Need the code? Click here</summary>

```{r bigmac_gg_code, eval=FALSE, echo=TRUE}
library(tidyverse)

# data processing 

## process the data file, using usd_adjusted as reference
## create a new variable for over/under

big_mac_processed = 
  big_mac %>% 
  filter(!is.na(usd_adjusted)) %>% 
  mutate(valued = if_else(usd_adjusted > 0, "over", "under")) 


# data visualization -- 1

gg_bigmac_turkey =
  big_mac_processed %>% 
  filter(name == "Turkey") %>% 
  
  ggplot(aes(x = date, y = usd_adjusted)) +
  
  # set the horizontal line first, as that will go under all the other geoms
  geom_hline(yintercept = 0, linetype = "longdash", alpha = 0.4)+
  
  geom_path(color = "darkred", size = 1)+
  geom_point(aes(shape = valued), color = "darkred", size = 3, stroke = 1, fill = "white") + 
  scale_shape_manual(values = c(21, 16))+
  
  labs(title = "The Big Mac Index: Turkey",
       subtitle = "The big mac index was invented by The Economist in 1986
       as a lighthearted guide to whether currencies are at their “correct” level.
       
       This graph shows how the Turkish Lira has compared to the US Dollar.",
       caption = "#TidyTuesday | Source: The Economist",
       x = "",
       y = "GDP adjusted index, 
       relative to the US dollar") + 
  
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 8, hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5),
    plot.margin = margin(t = 2, r = 1, b = 2, l = 1, unit = "cm")
  ) 


# data visualization -- 2

gg_bigmac_all = 
  big_mac_processed %>% 
  # select a few countries. this is a random list to get a 3x3 set of panels
  filter(name %in% c("Argentina", "Brazil", "Britain", 
                     "Canada", "Chile", "Egypt",
                     "Peru", "Poland", "Sweden")) %>% 
  
  ggplot(aes(x = date, y = usd_adjusted)) +
  geom_hline(yintercept = 0, linetype = "longdash", alpha = 0.4)+
  geom_path(color = "darkred")+
  geom_point(aes(shape = valued), color = "darkred", size = 1.5, fill = "white") + 
  scale_shape_manual(values = c(21, 16))+
  
  labs(title = "The Big Mac Index
       ",
       caption = "
       #TidyTuesday | Source: The Economist",
       x = "",
       y = "GDP adjusted index, 
       relative to the US dollar") + 
  facet_wrap(~name)+
  
  theme_void() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 8, hjust = 0.5),
    plot.caption = element_text(size = 8, hjust = 0.5),
    plot.margin = margin(t = 2, r = 1, b = 2, l = 1, unit = "cm")
  ) 
```
</details>


---

## Broadway Musicals
Source: [TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-04-28)

load data file here
```{r broadway_load, echo=T}
grosses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-28/grosses.csv', guess_max = 40000)
```

```{r broadway}
library(tidyverse)

total_grossing = 
  grosses %>% 
  group_by(show) %>% 
  dplyr::summarise(total_gross = sum(weekly_gross),
                   price_med = median(avg_ticket_price, na.rm = TRUE),
                   price_max = max(top_ticket_price, na.rm = TRUE)) %>% 
  arrange(total_gross) %>% 
  top_n(4, total_gross) %>% 
  mutate(total_gross_mill = total_gross/1000000,
         total_gross_bill = total_gross_mill/1000,
         total_gross_bill = round(total_gross_bill,1),
         label = paste0(show, ", $", total_gross_bill, " B"))

total_grossing %>% 
  ggplot()+
  geom_segment(aes(x = 1.18, xend = 1, 
                   y = total_gross_bill, yend = total_gross_bill),
               color = "grey30", linetype = "longdash", alpha = 0.5)+  
  geom_point(aes(x = (1), y = total_gross_bill, size = total_gross_bill), 
             alpha = 0.6, shape = 21, fill = "darkred", color = "darkred", stroke = 1.5,
  )+
  geom_text(aes(x = 1.19, y = total_gross_bill, label = label), hjust = 0, size = 4)+
  #geom_text(aes(x = 1.19, y = total_gross_bill-0.1, label = label), hjust = 0, size = 3)+
  
  labs(title = "Highest Grossing Broadway Shows",
       subtitle = "#TidyTuesday | Source: Playbill")+
  
  xlim(0.9, 1.5)+
  ylim(0.6, 2.0)+
  #scale_y_log10()+
  #geom_point(color = "darkred")+
  scale_size(range = c(30, 70))+
  #scale_size_area(max_size = 100)+
  theme_void()+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 8))


```

<details>
<summary>Need a hint? Click here</summary>

- This may look intimidating, but this is just a fancy-looking `geom_point`.
- Data are provided as weekly gross, you need to calculate the total gross per show, i.e.
`group_by(show) %>% summarize(...)`
- Then you choose the top 4 highest shows, using the `top_n()` function.
- Create a label and then use `geom_text`

</details>


<details>
<summary>Need the code? Click here</summary>
```{r broadway_code, eval=FALSE, echo=TRUE}
# first, process the data
# gross values are in terms of weekly gross, we need to calculate total gross per show
total_grossing = 
  grosses %>% 
  group_by(show) %>% 
  dplyr::summarise(total_gross = sum(weekly_gross)) %>% 
  arrange(total_gross) %>% 
  # filter only the top 4
  top_n(4, total_gross) %>%
  # we want to report the gross in terms of billion dollars
  # then create a `label` column that includes the show name and the total gross in $B
  mutate(total_gross_mill = total_gross/1000000,
         total_gross_bill = total_gross_mill/1000,
         total_gross_bill = round(total_gross_bill,1),
         label = paste0(show, ", $", total_gross_bill, " B"))

# now make the plot
total_grossing %>% 
  ggplot()+
  
  # make the segment first, so that goes to the back
  geom_segment(aes(x = 1.18, xend = 1, 
                   y = total_gross_bill, yend = total_gross_bill),
               color = "grey30", linetype = "longdash", alpha = 0.5)+  
  
  # because we want all the circles lined vertically, set x = constant value, e.g. 1
  geom_point(aes(x = 1, y = total_gross_bill, size = total_gross_bill), 
             alpha = 0.6, shape = 21, fill = "darkred", color = "darkred", stroke = 1.5)+
  
  # geom_text will add text labels to the plot, it works the same way as other geoms
  geom_text(aes(x = 1.19, y = total_gross_bill, label = label), hjust = 0, size = 4)+
  
  
  labs(title = "Highest Grossing Broadway Shows",
       subtitle = "#TidyTuesday | Source: Playbill")+
  
  xlim(0.9, 1.5)+
  ylim(0.6, 2.0)+

  scale_size(range = c(2, 70))+

  # set and customize theme 
  theme_void()+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 8))
```

</details>


## Transit Costs

load data file here
```{r transit_data}
library(tidyverse)
transit_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-01-05/transit_cost.csv')
```

```{r transit_graph}

transit_cost_longest = 
  transit_cost %>% 
  drop_na(country) %>% 
  top_n(10, length)


transit_cost_longest %>% 
  mutate(length = as.integer(length),
         line2 = paste0(line, " (", city, ")")) %>% 
  ggplot(aes(x = reorder(line2, -length), y = length, color = country))+
  
  geom_point(size = 10)+
  geom_segment(aes(xend = reorder(line2, -length), yend = 0), size = 5)+
  geom_text(aes(label = line2), 
            hjust = 1, angle = 90, size = 2.5,
            position = position_nudge(y = -5), color = "white")+
  geom_text(aes(label = length), 
            position = position_nudge(y = 15), fontface = "bold")+
  #scale_color_manual(values = c("blue", "green", "yellow", "orange", "red", "pink", "purple"))+
  scale_color_viridis_d()+
  labs(x = "",
       y = "",
       #title = "The 10 longest transit lines",
       #subtitle = "length in km",
       caption = "source = Transit Costs Project
       8 January 2021")+
  annotate("text", label = "The 10 longest transit lines 
           in the world",
           x = 10, y = 200, hjust = 1, fontface = "bold", size = 5)+
  annotate("text", label = "length in km", 
             angle = 90, x = 0.5, y = 100, size = 3, fontface = "italic")+ 
  
    theme_void()+
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 1),
    plot.caption = element_text(hjust = 0.5, color = "grey40", size = 8),
    legend.position = "none")

ggsave("transit.png", width = 5.5, height = 5.5)  

```

---
<details>
<summary>Session Info</summary>

**Kaizad F. Patel**

Date last run: `r Sys.Date()`

```{r}
sessionInfo()
```



</details>
