---
title: "ggplot2 -- working with themes"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```

```{r}
library(palmerpenguins)
library(ggplot2)
```

## Step 6: themes 
[*back to top*](#toc)

You can use pre-defined themes to change the look of your plots. 

The default is `theme_gray()`. Here are the available options.

```{r, echo=FALSE, fig.width=9, fig.height=4.5}
p = ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) + ylim(150, 250) + xlim(3000, 5000)

bw = p+theme_bw()+annotate("text", label = "theme_bw()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
classic = p+theme_classic()+annotate("text", label = "theme_classic()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
dark = p+theme_dark()+annotate("text", label = "theme_dark()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
gray = p+theme_gray()+annotate("text", label = "theme_gray()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
light = p+theme_light()+annotate("text", label = "theme_light()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
linedraw = p+theme_linedraw()+annotate("text", label = "theme_linedraw()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
minimal = void = p+theme_minimal()+annotate("text", label = "theme_minimal()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)
void = p+geom_point()+theme_void()+annotate("text", label = "theme_void()", 
                      x = 3000, y = 220, hjust = "left", vjust = "top", size = 3)


cowplot::plot_grid(bw, classic, dark, gray, light, linedraw, minimal, void, 
                   nrow = 2)
```

You can use `theme()` to customize other features of the plot, such as font size, font color, plot color, legend position, etc. 
[coming soon]

`element_blank()`

`element_text()`

`element_rect()`



---

<a name = "axis_text"></a>

#### axis titles and text

```{r}
p = ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point() + 
  theme_bw()
```

```{r}
p + theme(axis.title = element_text(face = "bold", 
                                    size = 14, 
                                    family = "Georgia"))
```

```{r}
p + theme(axis.title.x = element_text(face = "bold", 
                                    size = 14, 
                                    family = "Georgia"))
```

```{r}
p + theme(axis.text = element_text(size = 12, color = "red"))
```

---

<a name = "gridlines"></a>

##### removing gridlines
```{r}
p + theme(panel.grid = element_blank())
p + theme(panel.grid.major = element_blank())
```

---

<a name = "border"></a>

##### setting panel border
```{r}
p + theme(panel.border = element_rect(size = 2))
```

##### setting panel background
```{r}
p + theme(panel.background = element_rect(fill = "yellow"))
```

---

<a name = "format_title"></a>

##### adjusting title, subtitle
```{r}
p + labs(title = "Palmer Penguins", subtitle = "flipper length vs. body mass") +
  theme(plot.title = element_text(size = 22),
        plot.subtitle = element_text(size = 14, color = "darkred", family = "Georgia"))
```


---

---

