---
title: "Basic Univariate Statistics --  LME"
output: github_document
---

This tutorial will walk you through Linear Mixed Effects Models (LME).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

### What is an LME/LMM?

Linear mixed effects models (linear mixed models) allow us to use fixed as well as random variables in our linear model.
A random effect is something we want to control for, but not include as a primary explanatory effect (fixed).

This has practical applications for soil science/environmental science experiments where the same sample might be analyzed multiple times, e.g. before vs. after homogenization, before vs. after a treatment, or co-located split plots. In such cases, the treatment (homogenization, etc.) becomes the fixed variable, and the sample number becomes the random variable.

---

### Packages

The `{nlme}` and `{lme4}` packages are commonly used to perform LME.

We will use the `{nlme}` package for this tutorial. This package comes with a pre-loaded dataset, `Orthodont`

```{r}
library(nlme)
head(Orthodont)
```

---

### LME example

Each subject (M01-M16, F01-F13) was measured for orthodontic distance at different ages. You can get detailed info about the dataset using `?Orthodont` in your console.

Given the structure of this dataset, we would expect `distance` to vary with age and sex. However, each subject could also introduce variability, so we control for it by using `subject` as a random factor.

The LME setup is like a typical univariate model, y ~ x, but with the addition of the random effect

```{r}
l = lme(distance ~ age * Sex, random = ~1|Subject, data = Orthodont)
print(l)
summary(l)
```

To get p-values and determine if the fixed effects were significant, we use an ANOVA on the LME. There are two ways to do so, both give the same results. 

```{r, eval=FALSE}
anova(l)
```

```{r}
anova.lme(l)
```

#### Reporting these results

When reporting these results in a manuscript, etc., include the test name, the statistic, and the p-value.

For example: YY was significantly influenced by XX (LME, F = aaaa, P = bbbb)
