---
title: "dplyr basics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```

```{r}
library(tidyverse)
```

---

We will use the `{dplyr}` built-in dataset `starwars`

<details>
  <summary>click here to explore the `starwars` dataset</summary>

```{r}
names(starwars)
skimr::skim(starwars)

```

</details>


---

# `{dplyr}` functions - the basics

## rename

Use this to rename columns.  

general format: `rename(dataframe, "NEW_COLUMN_NAME" = "OLD_COLUMN_NAME")`

You can rename multiple columns in the same command.

```{r}
starwars_rn = rename(starwars, "NAME" = "name")
names(starwars_rn)

starwars_rn2 = rename(starwars, 
                      "NAME" = "name",
                      "HEIGHT" = "height",
                      "MASS" = "mass")
names(starwars_rn2)
```




```{r}
starwars %>% 
  rename("NAME" = "name") %>% 
  names()

starwars %>% 
  rename("NAME" = "name",
         "HEIGHT" = "height",
         "MASS" = "mass") %>% 
  names()
```

## select

Use this to subset certain columns. 

general format: `select()

You can do this multiple ways:

- choosing column names

```{r, eval=FALSE}
starwars %>% 
  select(name, height, mass) %>% 
  head()
```

- choosing column positions
```{r, eval=FALSE}
starwars %>% 
  select(1:3) %>% 
  head()
```

- deselecting column names/positions
```{r}
starwars %>% 
  select(-(4:14)) %>% 
  head()
```



## filter

Use `filter` to subset rows where certain conditions are met. 

For example, filter only characters with black hair 

```{r}
starwars %>% 
  filter(hair_color == "black") %>% 
  head()
```

#### multiple conditions 

For example, keep only characters with black or blond hair 
```{r}
starwars %>% 
  filter(hair_color %in% c("black", "blond")) %>% 
  head()
```
  

#### and/or conditions (multiple variables)
For example, keep only characters with black hair and blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" & eye_color == "blue") %>% 
  head()
```

keep only characters with black hair *or* blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" | eye_color == "blue") %>% 
  head()
```

#### "not" conditions

keep only characters without black hair 
```{r, eval=FALSE}
starwars %>% 
  filter(hair_color != "black") %>% 
  head()
```

another notation for this is 
```{r}
starwars %>% 
  filter(!hair_color == "black") %>% 
  head()
```

> **_NOTE:_**   
Note the symbols:  
== to meet a condition (for character variables only)  
| for "or" conditions  
! for "not"



## mutate

Use the `mutate()` function to either create new columns or modify existing columns.  
Basic usage is provided here, more advanced options are covered [here] and [here].


### create a new column


example: duplicating an existing column
```{r}
starwars %>% 
  dplyr::select(name) %>% 
  mutate(name2 = name) %>% 
  head() 
```


example: combining two columns
```{r}
starwars %>% 
  dplyr::select(name, gender) %>% 
  mutate(name_gen = paste(name, "-", gender)) %>% 
  head() 
```

Here, we use the `paste()` function to combine two columns and a hyphen.  
There are two versions of the `paste()` function:
- `paste()`: combines the pieces and adds a space between them
- `paste0()`: combines the pieces without adding a space

 

### modify an existing column


## summarize

## group_by
group operations



## arrange

## drop_na


---

<details>
  <summary>Session Info</summary>

Date Run: `r Sys.Date()`

```{r}
sessionInfo()
```

</details>