---
title: "using pipes"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```

### A `{dplyr}` example (without pipes)

```{r}
library(dplyr)
```


Consider this multi-part example: 

From the `starwars` dataset, you want to choose only characters with black hair. 

```{r}
sw_blackhair = filter(starwars, hair_color == "black")
sw_blackhair$name
names(sw_blackhair)
```

You then want to keep only columns related to name, height, and hair color

```{r}
sw_blackhair2 = select(sw_blackhair, name, height, hair_color)
names(sw_blackhair2)
```

You then want to rename the `hair_color` column

```{r}
sw_blackhair3 = rename(sw_blackhair2, "HairColor" = "hair_color")
names(sw_blackhair3)
```

### Pipes

In this example above, we created multiple files to get to our final desired output. An easier way to do this is using the **pipe** (%>%).

```{r}
sw2 = starwars %>% 
  filter(hair_color == "black") %>% 
  select(name, height, hair_color) %>% 
  rename("HairColor" = "hair_color")

names(sw2)
str(sw2)
```
