---
title: "dplyr basics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.height = 3, fig.width = 3)
```

```{r}
library(tidyverse)
```

---

We will use the `{dplyr}` built-in dataset `starwars`

<details>
  <summary>click here to explore the `starwars` dataset</summary>

```{r}
names(starwars)
skimr::skim(starwars)

```

</details>


---

# `{dplyr}` functions - the basics

---

## 1. rename

Use this to rename columns.  

general format: `rename(dataframe, "NEW_COLUMN_NAME" = "OLD_COLUMN_NAME")`

You can rename multiple columns in the same command.

```{r}
starwars_rn = rename(starwars, "NAME" = "name")
names(starwars_rn)

starwars_rn2 = rename(starwars, 
                      "NAME" = "name",
                      "HEIGHT" = "height",
                      "MASS" = "mass")
names(starwars_rn2)
```




```{r}
starwars %>% 
  rename("NAME" = "name") %>% 
  names()

starwars %>% 
  rename("NAME" = "name",
         "HEIGHT" = "height",
         "MASS" = "mass") %>% 
  names()
```


---

## 2. select

Use this to subset certain columns. 

general format: `select()`

You can do this multiple ways:

- choosing column names

```{r, eval=FALSE}
starwars %>% 
  select(name, height, mass) %>% 
  head()
```

- choosing column positions
```{r, eval=FALSE}
starwars %>% 
  select(1:3) %>% 
  head()
```

- deselecting column names/positions
```{r}
starwars %>% 
  select(-(4:14)) %>% 
  head()
```




---

## 3. filter

Use `filter` to subset rows where certain conditions are met. 

#### 3.1. single condition 

For example, filter only characters with black hair 

```{r}
starwars %>% 
  filter(hair_color == "black") %>% 
  head()
```

#### 3.2. multiple conditions 

For example, keep only characters with black or blond hair 
```{r}
starwars %>% 
  filter(hair_color %in% c("black", "blond")) %>% 
  head()
```
  

#### 3.3. and/or conditions (multiple variables)
For example, keep only characters with black hair and blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" & eye_color == "blue") %>% 
  head()
```

keep only characters with black hair *or* blue eyes 

```{r}
starwars %>% 
  filter(hair_color == "black" | eye_color == "blue") %>% 
  head()
```

#### 3.4. "not" conditions

keep only characters without black hair 
```{r, eval=FALSE}
starwars %>% 
  filter(hair_color != "black") %>% 
  head()
```

another notation for this is 
```{r}
starwars %>% 
  filter(!hair_color == "black") %>% 
  head()
```

> **_NOTE:_**   
Note the symbols:  
== to meet a condition (for character variables only)  
| for "or" conditions  
! for "not"


#### 3.5. `%in%` vs. `==`

By default, when using `==` or `!=`, the `filter()` command will also drop all NA values - even if we want to retain those rows.  
Therefore, a better alternative is to use `%in%`, as described above.

Examples:

Keep black hair color
```{r, eval = F}
starwars %>% 
  filter(hair_color %in% c("black")) %>% 
  head()
```

Remove black hair color
```{r, eval = F}
starwars %>% 
  filter(!hair_color %in% c("black")) %>% 
  head()
```

---

## 4. mutate

Use the `mutate()` function to either create new columns or modify existing columns.  
Basic usage is provided here, more advanced options are covered [here] and [here].


### 4.1. create a new column


example: duplicating an existing column
```{r}
starwars %>% 
  dplyr::select(name) %>% 
  mutate(name2 = name) %>% 
  head() 
```


example: combining two columns
```{r}
starwars %>% 
  dplyr::select(name, gender) %>% 
  mutate(name_gen = paste(name, "-", gender)) %>% 
  head() 
```

> Here, we use the `paste()` function to combine two columns and a hyphen.  
There are two versions of the `paste()` function:  
- `paste()`: combines the pieces and adds a space between them  
- `paste0()`: combines the pieces without adding a space

 

### 4.2. modify an existing column

Sometimes, you want to modify an existing column without creating a new column

Example: `height` is currently listed in cm, and we want to change this to m.

```{r}
starwars %>% 
#  dplyr::select(height) %>% 
  mutate(height = height/100) %>% 
  head() 
```

> **NOTE #1: Be very wary** when modifying an existing column, as it is very easy to make a mistake and not catch it.  
Try this example, what does it do? What happens if you run this same line of code two times?
```{r, eval = F}
starwars = starwars %>% dplyr::select(height) %>% mutate(height = height/100)
```

> **NOTE #2: UNITS.** Always include units in the column name, so there is no confusion about the data. The columns in the `starwars` do not include units (sigh), so we need to look up the [package/dataset info online](https://dplyr.tidyverse.org/reference/starwars.html) to fully understand the data. NOT IDEAL. Always, always, always include units.

Some other examples where you can use `mutate` to modify an existing column:

- changing the column type: `mutate(x = as.numeric(x))`, `mutate(y = as.character(y))`
- changing the order levels for a factor column: `mutate(x = factor(x, levels = c(X, Y, Z)))`
- rounding a numeric column to a certain number of decimal places: `mutate(x = round(x, 2))`

---

## 5. summarize

The `summarize` function allows you to calculate summary statistics (mean, median, standard deviation, etc.) for your data

```{r, eval=FALSE}
starwars %>% 
  summarise(height_mean = mean(height),
            height_median = median(height),
            height_sd = sd(height))
```

This line of code above gives us NA in the output - because there are some NA values in the `height` column.  
To avoid this issue, include `na.rm = TRUE` in the arguments, to remove the NAs.

```{r}
starwars %>% 
  summarise(height_mean = mean(height, na.rm = TRUE),
            height_median = median(height, na.rm = TRUE),
            height_sd = sd(height, na.rm = TRUE))
```

## 6. group_by
group operations




You can use `group_by()` with `mutate` and `summarize` to perform the calculations by group, i.e. to get summary statistics, etc. for each set of values, given a grouping column.

Example: We previously calculated mean height for the entire dataset. But what if we want mean height for each species?

```{r}
starwars %>% 
  group_by(species) %>% 
  dplyr::summarise(height_mean = mean(height, na.rm = TRUE),
                  n = n())
```

The `n()` function used here gives us the count, or the number of rows present in that group. So, there was only 1 from the Aleena species, but 6 Droids.

> **NOTE:** because of package masking, sometimes summarize does not work well with group_by(), giving us a single value instead of grouped. The workaround for this is using the namespace (`dplyr::summarize()`)


---

<details>
  <summary>Session Info</summary>

Date Run: `r Sys.Date()`

```{r}
sessionInfo()
```

</details>